---
output: github_document
---

<!-- README.md is generated from README.Rmd. Please edit that file -->

```{r, echo = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  fig.path = "readme_figures/README-"
)
```

# NetworkInference: Inferring Latent Diffusion Networks

![](https://travis-ci.org/desmarais-lab/NetworkInference.svg)

## About

This package provides an R implementation of the netinf algorithm 
created by Gomez-Rodriguez, Leskovec, and Krause (see 
[here](http://snap.stanford.edu/netinf/) for more information and the original 
C++ implementation). Given a set of events that spread between a set of nodes 
the algorithm infers the most likely stable diffusion network that is underlying 
the diffusion process.

## Installation

The package can be installed from [CRAN](https://CRAN.R-project.org/):
```{r, eval=FALSE}
install.packages("NetworkInference")
```

The latest development version can be installed from 
[github](https://github.com/desmarais-lab/NetworkInference):
```{r, eval=FALSE}
#install.packages(devtools)
devtools::install_github('desmarais-lab/NetworkInference')
```

## Quick start guide

This gets you started. For more in-depth explanations see the package vignette 
and documentation.

First, get your data into the `cascades` format required by the `netinf` 
function:

```{r, results='hide', message=FALSE}
library(NetworkInference)

# Simulate random cascade data
set.seed(423399)
df <- simulate_rnd_cascades(50, n_node = 26) # Simulates random cascade data
node_names <- unique(df$node_name)

# Cast data into `cascades` object
## From dataframe
cascades <- as.cascade(df, node_names = node_names)

## From matrix
df_matrix <- as.matrix(cascades) ### Create example matrix
cascades <- as.cascade(df_matrix, node_names = node_names)
```

Cascades can be explored using `summary` and `plot`. See the vignette and 
package documentation for full functionality.
```{r, fig.align='center', fig.width=7, fig.height=4}
cascade_ids <- names(cascades[[1]])
selection <- cascade_ids[c(1, 10)]
plot(cascades, label_nodes = TRUE, selection = selection)
selection <- cascade_ids[c(1:20)]
plot(cascades, label_nodes = FALSE, selection = selection)
```

The actual netinf algorithm is used with:
```{r}
result <- netinf(cascades, trans_mod = "exponential", lambda = 1, n_edges = 5)
```

The resulting network comes in form of an edgelist:
```{r, eval=FALSE}
print(result)
```
```{r, results="asis", echo=FALSE}
pander::pandoc.table(result)
```

The network as well as the improvement in fit by each edge can be visualized:
```{r}
plot(result, type = "network")
plot(result, type = "improvement")
```
