---
title: "NetworkInference: Inferring Latent Diffusion Networks"
author: "Fridolin Linder"
date: "`r Sys.Date()`"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{NetworkInference: Inferring Latent Diffusion Networks}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

---

# Introduction

---

This package provides an R implementation of the netinf algorithm 
created by @gomez2010inferring (see [here](http://snap.stanford.edu/netinf/) 
for more information and the original C++ implementation). Given a set of events 
that spread between a set of nodes the algorithm infers the most likely stable 
diffusion network that is underlying the diffusion process.

---

# Installation

---

The package can be installed from [CRAN](https://CRAN.R-project.org/):
```{r, eval=FALSE}
install.packages("NetworkInference")
```

The latest development version can be installed from 
[github](https://github.com/desmarais-lab/NetworkInference):
```{r, eval=FALSE}
#install.packages(devtools)
devtools::install_github('desmarais-lab/NetworkInference')
```

---

# Quick start guide

---

To get started, get your data into the `cascades` format required by the `netinf` 
function:

```{r, results='hide', message=FALSE}
library(NetworkInference)

# Simulate random cascade data
df <- simulate_rnd_cascades(50, n_node = 20)

# Cast data into `cascades` object
## From long format
cascades <- as_cascade_long(df)

## From wide format
df_matrix <- as.matrix(cascades) ### Create example matrix
cascades <- as_cascade_wide(df_matrix)
```

Then fit the model:
```{r}
result <- netinf(cascades, quiet = TRUE)
```

```{r, eval=FALSE}
head(result)
```
```{r, results="asis", echo=FALSE}
pander::pandoc.table(head(result))
```

---

# Tutorial

---

This is a quick tutorial to get started with the package. For more detailed 
information on the algorithm and functionality of the package see the 
package documentation.

`netinf` infers the optimal diffusion network from a set of *nodes* 
and a number of so called *cascades*. A cascade is a series of events occurring at 
a specified time. For this demo we will replicate the analysis presented in 
@desmarais2015persistent including additional data collected in the [SPID Database](https://doi.org/10.7910/DVN/CVYSR7). In this paper Desmarais et al. infer a latent network
for policy diffusion based on adoption of policies in the US states. In this
case a node in the network is a state, a cascade refers to a specific policy and
an event is the adoption of said policy in a state.  

## Preparing the Data

The policy adoption data is available in the package:

```{r, message=FALSE}
library(NetworkInference)
# Load the `policies` dataset (?policies for details).
data(policies)
```

The data is stored as a `data.frame` where each row corresponds to an adoption event.

```{r, eval=FALSE}
head(policies)
```

```{r, results="asis", echo=FALSE}
pander::pandoc.table(head(policies))
```

The first column (`statenam`) gives the state that adopted the `policy` in year `adopt_year`. More information on the policies can be found in the [SPID repository](https://doi.org/10.7910/DVN/CVYSR7).

Most functionality of the `NetworkInference` package is based on the `cascades` 
data format. So before starting with the analysis we have to transform our data
to such an object.

```{r}
policy_cascades <- as_cascade_long(policies, cascade_node_name = 'statenam',
                                   event_time = 'adopt_year', 
                                   cascade_id = 'policy')
```

In this case we used the function `as_cascade_long`. If your data is in wide format you can convert it using the function `as_cascade_wide`.

### Visually Inspecting Cascades

It's always good practice to visually inspect the data before working with it. 
The `NetworkInference` package provides functionality to visualize the cascade 
data. 

The function `summary.cascades()` provides quick summary statistics on the 
cascade data:

```{r}
summary(policy_cascades)
```

The first four lines provide the number of cascades, the number of nodes in the 
system, the number of nodes involved in cascades (there might be states that we 
don't have diffusion data on, but we still want them represented in the dataset)
and the possible number of edges in a potential diffusion network (a diffusion edge 
between nodes `u` and `v` only makes sense if there is at least one cascade in 
which `u` experiences an event before `v`). In this example there are 187 
policies and 50 states. Each state is involved in at least one policy cascade 
and a fully connected diffusion network would have 2450 edges. 

It also provides summary statistics on the distribution of the cascade lengths 
(number of nodes involved in each cascade) and the number of ties in the 
cascades (two nodes experiencing the same event at the same time). For our 
example, we can see that the 'smallest' policy was adopted by 10 states and the
'largest' by all 50 states. From the tie summaries we see that there is at least
one policy that was adopted by 45 states in the same year. 

The `plot()` method allows to plot cascades with varying degrees of detail. The 
argument `label_nodes` (`TRUE/FALSE`) provides node labels which require more space
but provide more detail. The argument `selection` allows to pick a subset of 
cascades to visualize in case there are too many to plot. If `label_nodes` is set 
to `FALSE` each event is depicted by a dot, which allows to visualize more cascades
simultaneously. 

Let's first look at the visualization with labels. Here we plot two cascades, 
selected by their name:
```{r, fig.align='center', fig.width=7, fig.height=4}
selection <- c('guncontrol_assaultweapon_ba', 'guncontrol_licenses_dealer')
plot(policy_cascades, label_nodes = TRUE, selection = selection)
```

We can also plot more cascades with less detail:
```{r, fig.align='center', fig.width=7, fig.height=4}
selection <- c('waiting', 'threestrikes', 'unionlimits', 'smokeban', 
               'paperterror', 'miglab', 'methpre', 'lott', 'lemon', 'idtheft',
               'harass', 'hatecrime', 'equalpay')
plot(policy_cascades, label_nodes = FALSE, selection = selection)
```

This produces a ['violin plot'](https://en.wikipedia.org/wiki/Violin_plot) for each cascade with the single diffusion events overplotted as dots. As we already saw in the previous visualization, the policy data has a lot of ties (i.e. many states adopted a policy in the same year) which is indicated by the areas of higher density in the violin plot.


## Inferring the Latent Diffusion Network

The `netinf` algorithm is implemented in the `netinf()` function. The `netinf` inferrs edges based on a diffusion model. That is we assume a parametric model for the diffusion times between edges. Currently three different diffusion models are implemented: The exponential distribution, the rayleigh distribution and the log-normal distribution. The model can be chosen with the `trans_mod` argument (default is the exponential distribution). 

### Classic Netinf

In the original implementation the number of edges to infer had to be fixed and chosen by the researcher. If you want to run `netinf` in this classic way you can do so by specifiying all parameters and the number of edges:

```{r}
results <- netinf(policy_cascades, trans_mod = "exponential", n_edges = 100, 
                  params = 0.5, quiet = TRUE)
```

The exponential model has only one parameter (lambda or the rate). If there are more parameters the details section in the documentation of the `netinf` function (`?netinf`) has more detail on how to specify parameters and on the specific parametrization used by the implementation. 

`n_edges` specifies how many edges should be inferred. See @gomez2010inferring and @desmarais2015persistent for guidance on choosing this parameter if running netinf in classic mode. If the number of edges is specified manually, it has to be lower than the maximum number of possible edges. An edge `u->v` is only possible if in at least one cascade `u` experiences an event *before* `v`. This means, that the maximum number of edges depends on the data. The function `count_possible_edges()` allows us to compute the maximum number of edges in advance:
 
```{r}
npe <- count_possible_edges(policy_cascades)
npe
```


### Automatic Parameter Selection

With version 1.2.0 `netinf` can be run without manually specifying the number of edges or the parameters of the diffusion model. 

#### Selecting the number of edges automatically

After each iteration of the netinf algorithm, we check if the edge added significant improvement to the network. This is done via a vuong style test. Given the likelihood score for each cascade conditional on the network inferred so far, we penalize the network with one addional edge and test if the increase in likelihood accross all cascades is significant. The user still has to specify a p-value cut-off. If the p-value of an edge is larger than the specified cut-off the algorithm stops inferring more edges. The cut-off is again set via the `n_edges` argument. Setting it to a value in the interval (0,1) let's the funciton use the p-value procedure and uses that cut-off:

```{r}
results <- netinf(policy_cascades, trans_mod = "exponential", n_edges = 0.1, 
                  params = 0.5, quiet = TRUE)
nrow(results)
```

We see that with a fixed lambda of 0.5 and a p-value cut-off of 0.1 the algorithm inferred 872 edges.

#### Selecting the parameters of the diffusion model

The diffusion model parameters can be selected automatically. Setting the `params` argument to `NULL` (default value) makes the `netinf` function initialize the parameters automatically. The parameters are initialized at the midpoint between the MLE of the minimum diffusion times and the MLE of the maximum diffusion times, across all cascades. Edges are then inferred until either the p-value cut-off or a manually specified number of edges (`n_edges`) is reached. 

```{r}
results <- netinf(policy_cascades, trans_mod = "exponential", n_edges = 0.1, 
                  , quiet = TRUE)
nrow(results)
```

### Netinf output

Let's take a look at the output of the algorithm. The output is a dataframe containing the inferred latent network in the form of an edgelist:

```{r, eval=FALSE, echo=TRUE}
head(results)
```
```{r, results = "asis", echo=FALSE}
pander::pandoc.table(head(results))
```

Each row corresponds to a directed edge. The first column indicates the origin node, the second the destination node. The third column displays the gain in model fit from each added edge. The last column displays the p-value from the Vuong test of each edge. There is a generic plot method to inspect the results. If more tweaking is required, the results are a dataframe so it should be easy for the more experienced users to make your own plot. With `type = "improvement"` the improvement from each edge can be plotted:

```{r, fig.align='center', fig.width=7, fig.height=4}
plot(results, type = "improvement")
```

We can also quickly check the p-value from the Vuong test associated with each edge addition:

```{r, fig.align='center', fig.width=7, fig.height=4}
plot(results, type = 'p-value')
```

In order to produce a quick visualization of the resulting diffusion network we can use the plot method again, this time with `type = "network"`. Note that in order to use this functionality the igraph package has to be installed.

```{r, fig.width=7, fig.height=5.5}
#install.packages('igraph')
# For this functionality the igraph package has to be installed
# This code is only executed if the package is found:
if(requireNamespace("igraph", quietly = TRUE)) {
    plot(results, type = "network")
}
```

If additional tweaking of the plot is desired, the network can be visualized using `igraph` explicitly. We refer you you to the [igraph documentation](https://CRAN.R-project.org/package=igraph) for details on how to customize the plot.

```{r, message=FALSE, eval=FALSE}
if(requireNamespace("igraph", quietly = TRUE)) {
    library(igraph) 
    g <- graph_from_data_frame(d = results[, 1:2])
    plot(g, edge.arrow.size=.3, vertex.color = "grey70")
}
```
